Log or not - Analysis of ECDC forecasts

### 2022-01-26 analysis/Analysis ECDC forecasts.Rmd
generate some plots for Hub evaluation

hub_data <- rbindlist(list(
  fread(here("data", "full-data-european-forecast-hub-1.csv")), 
  fread(here("data", "full-data-european-forecast-hub-2.csv"))
)) |>
  unique()

@@

scores <- bind_rows(scores_natural, scores_log)
fwrite(scores, here("output", "data", "all-scores-european-hub.csv"))
```

```{r}
scores <- fread(here("output", "data", "all-scores-european-hub.csv"))
```

@@

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  ggplot(aes(y = interval_score, x = target_type)) + 
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) + 
  geom_boxplot(alpha = 0.5) + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") + 
  labs(y = "Interval score", x = "Target type")
```

@@

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(target_type, location, scale) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  ggplot(aes(y = interval_score, x = reorder(location, -interval_score))) + 
  geom_bar(stat = "identity") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ target_type + scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn) + 
  labs(y = "Interval score", x = "Target type")
```

@@

@@

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location == "GB",
         horizon == 2) |>
  group_by(target_type, scale) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")
```

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location == "FR",
         horizon == 2) |>
  group_by(target_type, scale) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")
```

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location == "ES",
         horizon == 2) |>
  group_by(target_type, scale) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")
```

@@

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble") |>
  group_by(scale, target_type, horizon) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = horizon)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale) + 
  theme_minimal() + 
  labs(x = "Forecast horizon in weeks", y = "Multplicative change in interval score")
```

@@

--> not sure that is useful

@@

# check normality of residuals
check_normality(reg_log) |>
  plot()

@@

>>>>

hub_data <- rbindlist(list(
  fread(here("data", "full-data-european-forecast-hub-1.csv")), 
  fread(here("data", "full-data-european-forecast-hub-2.csv"))
)) |>
  unique()
```

```{r}
# create version of data where the true value is the true growth rate
# interesting question: does it make a difference, whether we transform it 
# to be the weekly growth rate, i.e. take the n-th root, where n is the horizon?

latest_truth <- fread(here("data", "weekly-truth-Europe.csv")) |>
  rename(last_known_true_value = true_value) |>
  mutate(forecast_date = target_end_date + 2) |>
  select(location, target_type, forecast_date, last_known_true_value) |>
  unique()

# unsure whether the pmax hack is the right way to do it
growth_data <- hub_data |>
  inner_join(latest_truth) |>
  mutate(true_value = true_value / pmax(last_known_true_value + 1),
         prediction = prediction / pmax(last_known_true_value + 1)) |>
  select(-last_known_true_value)
```

@@

scores_gr_natural <- score_forecasts(growth_data)
scores_gr_log <- score_forecasts(growth_data, scale = "log")

scores <- bind_rows(scores_natural, scores_log)
fwrite(scores, here("output", "data", "all-scores-european-hub.csv"))

scores_gr <- bind_rows(scores_gr_natural, scores_gr_log)
fwrite(scores_gr, here("output", "data", "all-scores-growth-rate-european-hub.csv"))
```

```{r}
scores <- fread(here("output", "data", "all-scores-european-hub.csv"))
scores[, scale := factor(scale, levels = c("natural", "log"))]

scores_gr <- fread(here("output", "data", "all-scores-growth-rate-european-hub.csv"))
scores_gr[, scale := factor(scale, levels = c("natural", "log"))]
```

@@

This uses the Hub ensemble

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  ggplot(aes(y = interval_score, x = target_type)) + 
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) + 
  geom_boxplot(alpha = 0.5) + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") + 
  labs(y = "Interval score", x = "Target type")

ggsave("output/figures/HUB-average-scores.png", width = 7, height = 5)
```

@@

```{r}
# alternatively, we could think about just summarising across all models for these plots? 
scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(target_type, location, scale) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  ggplot(aes(y = interval_score, x = reorder(location, -interval_score))) + 
  geom_bar(stat = "identity") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ target_type + scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn) + 
  labs(y = "Interval score", x = "Target type")

ggsave("output/figures/HUB-scores-locations.png", width = 7, height = 5)
```

@@

## Variance scores across locations

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(target_type, location, scale) |>
  summarise(mean_wis = mean(interval_score), 
            sd_wis = sd(interval_score), 
            .groups = "drop_last") |>
  ggplot(aes(y = sd_wis, x = mean_wis)) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", size = 0.5) + 
  geom_point() + 
  facet_wrap(~ target_type + scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  # scale_x_continuous(labels = scale_fn) + 
  # scale_y_continuous(labels = scale_fn) + 
  scale_x_continuous(labels = scale_fn, trans = "log10") +
  scale_y_continuous(labels = scale_fn, trans = "log10") +
  labs(x = "Mean WIS", y = "Sd of WIS")

ggsave("output/figures/HUB-sd-vs-mean-scores.png", width = 7, height = 5)

@@

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location %in% c("GB", "ES", "DE"),
         horizon == 2) |>
  rename(Location = location) |>
  group_by(target_type, scale, Location) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date, colour = Location)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")

ggsave("output/figures/HUB-scores-over-time.png", width = 7, height = 5)
```

@@

```{r}
scores |> 
  filter(model == "EuroCOVIDhub-ensemble") |>
  group_by(scale, target_type, horizon) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = horizon)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale) + 
  theme_minimal() + 
  labs(x = "Forecast horizon in weeks", y = "Multplicative change in interval score")

ggsave("output/figures/HUB-scores-over-horizon.png", width = 7, height = 5)

@@

@@

# check normality of residuals
check_normality(reg_log) |>
  plot(type = "qq")

@@

---

# Looking at growth rates directly

## Scores of the growth rate across targets

This uses the Hub ensemble

```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  ggplot(aes(y = interval_score, x = target_type)) + 
  geom_violin(aes(fill = target_type), alpha = 0.2, color = NA) + 
  geom_boxplot(alpha = 0.5) + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn, trans = "log10") + 
  labs(y = "Interval score", x = "Target type")
```

*Scoring the growth rate makes things more comparable*

Mean scores: 
```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(scale, target_type) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  make_table()
```

## Scores for the growth rate across locations

```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble", 
         horizon == 2) |>
  group_by(target_type, location, scale) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  ggplot(aes(y = interval_score, x = reorder(location, -interval_score))) + 
  geom_bar(stat = "identity") + 
  scale_fill_brewer(palette = "Set1", name = "Forecast target") + 
  facet_wrap(~ target_type + scale, scale = "free") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = scale_fn) + 
  labs(y = "Interval score", x = "Target type")
```

## Scores for growth rate over time

```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location == "GB",
         horizon == 2) |>
  group_by(target_type, scale) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")
```

```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location == "FR",
         horizon == 2) |>
  group_by(target_type, scale) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")
```

```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble",
         location == "ES",
         horizon == 2) |>
  group_by(target_type, scale) |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = forecast_date)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale, scale = "free_y") + 
  theme_minimal() + 
  labs(x = "Forecast date", y = "WIS relative to smallest observed value")
```

## Change of average scores for the growth rate for increasing forecast horizons

```{r}
scores_gr |> 
  filter(model == "EuroCOVIDhub-ensemble") |>
  group_by(scale, target_type, horizon) |>
  summarise(interval_score = mean(interval_score), 
            .groups = "drop_last") |>
  mutate(multiplicative_score = interval_score / min(interval_score)) |>
  ggplot(aes(y = multiplicative_score, x = horizon)) + 
  geom_line() + 
  geom_point() + 
  facet_grid(target_type ~ scale) + 
  theme_minimal() + 
  labs(x = "Forecast horizon in weeks", y = "Multplicative change in interval score")
```

## Distribution of scores growth rate

```{r}
scores_gr |>
  ggplot(aes(x = interval_score)) + 
  geom_density() + 
  facet_wrap(~ target_type + scale, scale = "free") + 
  scale_x_continuous(labels = scale_fn)
```

## Fit a model to growth rate data

```{r echo=TRUE}
reg_gr <- scores_gr |>
  filter(scale == "natural") |>
  glm(formula = interval_score ~ 1 + model + location + target_type + horizon, 
      family = gaussian)

reg_gr_log <- scores_gr |>
  filter(scale == "log") |>
  glm(formula = interval_score ~ 1 + model + location + target_type + horizon, 
      family = gaussian)
```

```{r}
# check distribution
performance::check_distribution(reg_gr_log) |>
  plot()

# check normality of residuals
check_normality(reg_gr_log) |>
  plot(type = "qq")

# heteroskedasticity of error terms - the plot takes extremly long to run
performance::check_heteroskedasticity(reg_gr_log)
# performance::check_heteroskedasticity(reg_log) |>
#   plot()

# check influence of outliers - plot also takes very long to run
performance::check_outliers(reg_gr_log)
# performance::check_outliers(reg_log) |>
#   plot()
